---
title: "Report for Choose Your Own project submission Project HarvardX PH125.9x- Heart Disease Risk"
output:  pdf_document
date: "2022-12-18"
---



## Introduction
In this report i decided to implement  Machine Learning Ensemble to Make Heart Disease Prediction. In Mongolia where i live, heart disease is number one death cause and Mongolia is ranked 9th in the World of heart disease death rate per 100000 as stated in World Health Rankings website https://www.worldlifeexpectancy.com/cause-of-death/coronary-heart-disease/by-country/. Since we do not have available dataset of our country i used  heart disease data set from University of California Irvine machine learning repository. This data set consist fo 14 different features and 303 observations. The description of the features from the website is the following:. 
*age: age in years
*sex: sex (1 = male; 0 = female)
*cp: chest pain type
Value 1: typical angina
Value 2: atypical angina
Value 3: non-anginal pain
*trestbps: resting blood pressure (in mm Hg on admission to the hospital)
*chol: serum cholestoral in mg/dl
*fbs: (fasting blood sugar > 120 mg/dl) (1 = true; 0 = false)
*restecg: resting electrocardiographic results
Value 0: normal
Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
Value 2: showing probable or definite left ventricular hypertrophy by Estes' criteria
*thalach: maximum heart rate achieved
*exang: exercise induced angina (1 = yes; 0 = no)
*oldpeak = ST depression induced by exercise relative to rest
*slope: the slope of the peak exercise ST segment
Value 1: upsloping
Value 2: flat
Value 3: downsloping
*ca: number of major vessels (0-3) colored by flourosopy
*thal: 3 = normal; 6 = fixed defect; 7 = reversable defect
*target: diagnosis of heart disease (angiographic disease status)
Value 0: < 50% diameter narrowing
Value 1: > 50% diameter narrowing
The target feature is the feature we will be trying to predict for this project.

## Load libraries  
* tidyverse: For data cleaning, sorting, and visualization  
* DataExplorer: For Exploratory Data Analysis  
* gridExtra: To plot several plots in one figure  
* ggpubr: To prepare publication-ready plots  
* GGally: For correlations  
* caTools: For classification model 
* rpart: For classification model  
* rattle: Plot nicer descision trees  
* randomForest: For Random Forest model
* library(caret)
* library(dplyr)
* library(matrixStats)
* library(gam)
* library(evtree)
* library(knitr)

```{r message = FALSE, warning = FALSE}
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(caret)) install.packages("dplyr")
if(!require(dplyr)) install.packages("dplyr")
if(!require(matrixStats)) install.packages("matrixStats")
if(!require(gam)) install.packages("gam")
if(!require(evtree)) install.packages("evtree")
if(!require(knitr)) install.packages("knitr")

library(tidyverse)# For data cleaning, sorting, and visualization
library(caret)
library(dplyr)
library(matrixStats)
library(gam)
library(evtree)
library(knitr)
library(DataExplorer) # For Exploratory Data Analysis
library(gridExtra) # To plot several plots in one figure
library(ggpubr) # To prepare publication-ready plots
library(GGally) # For correlations
library(caTools) # For classification model
library(rpart) # For classification model
library(rattle) # Plot nicer descision trees
library(randomForest) # For Random Forest model
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE,
	paged.print=TRUE
)
```

## I Data


Link to the UCI heart disease data: https://archive.ics.uci.edu/ml/datasets/heart+disease
Kaggle heart dataset: https://www.kaggle.com/datasets/zhaoyingzhu/heartcsv

**As explained on the links above, it is essential to note that on this dataset, the target value 0 indicates that the patient has heart disease.**

Attribute Information:  
age: age in years  
sex: (1 = male; 0 = female)  
cp: chest pain type  (typical angina, atypical angina, non-angina, or asymptomatic angina)  
trestbps: resting blood pressure (in mm Hg on admission to the hospital)  
chol: serum cholestoral in mg/dl  
fbs: Fasting blood sugar (< 120 mg/dl or > 120 mg/dl) (1 = true; 0 = false)  
restecg: resting electrocardiographic results (normal, ST-T wave abnormality, or left ventricular hypertrophy)  
thalach: Max. heart rate achieved during thalium stress test  
exang: Exercise induced angina (1 = yes; 0 = no)  
oldpeak: ST depression induced by exercise relative to rest  
slope: Slope of peak exercise ST segment (0 = upsloping, 1 = flat, or 2 = downsloping)  
ca: number of major vessels (0-3) colored by flourosopy  4 = NA  
thal: Thalium stress test result 3 = normal; 6 = fixed defect; 7 = reversable defect  0 = NA  
target: Heart disease status 1 or 0 (0 = heart disease 1 = asymptomatic) 

```{r message = FALSE}
heart <- read.csv("C:/Users/Hitech/Downloads/Heart.csv")
```
## Methods & Analysis

To achieve this goal we will be creating 9 different models and an ensemble and comparing their results. Because the nature of the problem is to determine if a patient is negative or positive, i.e 0 or 1, this is a binary classification problem and we will pick 10 algorithms that work well with binary classification. The algorithms we will be using are the following:

### Data Preparation

```{r}
copy <- heart

heart2 <- heart %>%
  filter(
    thal != 0 & ca != 4 # remove values correspondind to NA in original dataset
  ) %>%
# Recode the categorical variables as factors using the dplyr library.
  mutate(
    sex = case_when(
      sex == 0 ~ "female",
      sex == 1 ~ "male"
           ),
    fbs = case_when(
      fbs == 0 ~ "<=120",
      fbs == 1 ~ ">120"
            ),
    exang = case_when(
      exang == 0 ~ "no",
      exang == 1 ~ "yes"
            ),
    cp = case_when(
      cp == 3 ~ "typical angina",
      cp == 1 ~ "atypical angina",
      cp == 2 ~ "non-anginal",
      cp == 0 ~ "asymptomatic angina"
          ),
    restecg = case_when(
      restecg == 0 ~ "hypertrophy",
      restecg == 1 ~ "normal",
      restecg == 2 ~ "wave abnormality"
              ),
    target = case_when(
      target == 1 ~ "asymptomatic",
      target == 0 ~ "heart-disease"
              ),
    slope = case_when(
      slope == 2 ~ "upsloping",
      slope == 1 ~ "flat",
      slope == 0 ~ "downsloping"
    ),
    thal = case_when(
      thal == 1 ~ "fixed defect",
      thal == 2 ~ "normal",
      thal == 3 ~ "reversable defect"
    ),
    sex = as.factor(sex),
    fbs = as.factor(fbs),
    exang = as.factor(exang),
    cp = as.factor(cp),
    slope = as.factor(slope),
    ca = as.factor(ca),
    thal = as.factor(thal)
  )

glimpse(heart2) # Check that the transformnation worked
plot_missing(heart2) # Check that the transformation did not induce NA values
heart <- heart2 # Replace the heart dataset by the tidy dataset
```

## Data exploration


```{r}
heart %>%
  summary()
```

Use the DataExplorer library to get a sense of the distribution of the continuous and categorical variables.

```{r}
plot_density(heart, ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
plot_bar(heart, ggtheme = theme_classic2())
```

The next step is to combine dplyr and Data Explorer libraries to visualize the variables according to gender and disease.  

### B Analyze each variable per gender

```{r}
heart %>%
  filter(sex == "female") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
heart %>%
  filter(sex == "male") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
```

```{r}
heart %>%
  filter(sex == "female") %>%
  plot_bar(ggtheme = theme_classic2())
heart %>%
  filter(sex == "male") %>%
  plot_bar(ggtheme = theme_classic2())
```

### C Visualize variables per disease status

```{r}
heart %>%
  filter(target == "asymptomatic") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
heart %>%
  filter(target == "heart-disease") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
```

```{r}
heart %>%
  filter(target == "asymptomatic") %>%
  plot_bar(ggtheme = theme_classic2())
heart %>%
  filter(target == "heart-disease") %>%
  plot_bar(ggtheme = theme_classic2())
```

### Visualize the data per gender and disease status
```{r}
heart %>%
  filter(sex == "female", target == "asymptomatic") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
heart %>%
  filter(sex == "female", target == "heart-disease") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
```

```{r}
heart %>%
  filter(sex == "female", target == "asymptomatic") %>%
  plot_bar(ggtheme = theme_classic2())
heart %>%
  filter(sex == "female", target == "heart-disease") %>%
  plot_bar(ggtheme = theme_classic2())
```

```{r}
heart %>%
  filter(sex == "male", target == "asymptomatic") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
heart %>%
  filter(sex == "male", target == "heart-disease") %>%
  plot_density(ggtheme = theme_classic2(), geom_density_args = list("fill" = "black", "alpha" = 0.6))
```

```{r}
heart %>%
  filter(sex == "male", target == "asymptomatic") %>%
  plot_bar(ggtheme = theme_classic2())
heart %>%
  filter(sex == "male", target == "heart-disease") %>%
  plot_bar(ggtheme = theme_classic2())
```

### Prepare a summary table per disease and gender
```{r}
heart %>%
  group_by(target, sex) %>%
  summarise(
    n_disease = n(),
    mean_age = round(mean(age), digits=2),
    sd_age = round(sd(age), digits=2),
    mean_trestbps = round(mean(trestbps), digits=2),
    sd_trestbps = round(sd(trestbps), digits=2),
    mean_chol = round(mean(chol), digits=2),
    sd_chol = round(sd(chol), digits=2),
    mean_thalach = round(mean(thalach), digits=2),
    sd_thalach = round(sd(thalach), digits=2),
    mean_oldpeak = round(mean(oldpeak), digits=2),
    sd_oldpeak = round(sd(oldpeak), digits=2)
  )
```

## Visualization

From the Exploratory Data analysis, it seems that several differences are statistically significant according to gender and health status. 

### A Visualization of variables per gender

```{r}
# Male and Female count
a1 <- ggplot(heart, aes(x = sex, fill = sex)) +
  geom_bar(width = 0.5) + 
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# Age per gender
b1 <- ggplot(heart, aes(x= sex, y = age, fill = sex)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) +
  ylim(0, 90) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# trestbps
c1 <- ggplot(heart, aes(x = sex, y = trestbps, fill = sex)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "blood pressure (mmHg)") +
  ylim(0,250) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# chol
d1 <- ggplot(heart, aes(x = sex, y = chol, fill = sex)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "cholestorol (mg/dl)") +
  ylim(0,500) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# oldpeak
e1 <- ggplot(heart, aes(x = sex, y = oldpeak, fill = sex)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "ST depression") +
  ylim(0,10) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# thalach
f1 <- ggplot(heart, aes(x = sex, y = thalach, fill = sex)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "Max. heart rate") +
  ylim(0,250) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

suppressWarnings(ggarrange(a1, b1, c1, d1, e1, f1, 
          ncol = 2, nrow = 3,
            align = "v"))
```

```{r}
# Disease status
g1 <- ggplot(heart, aes(x = target, fill = sex)) +
  geom_bar(width = 0.5, position = 'dodge') + 
  labs(x = "") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# cp
h1 <- ggplot(heart, aes(cp, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "chest pain") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# restecg
i1 <- ggplot(heart, aes(restecg, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "rest. electrocardiographic") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# slope
j1 <- ggplot(heart, aes(slope, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "peak exercise ST") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# thal 
k1 <- ggplot(heart, aes(thal, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Thalium stress test") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# fbp
l1 <- ggplot(heart, aes(fbs, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Fasting blood sugar") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# exang
m1 <- ggplot(heart, aes(exang, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Exercise induced angina") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

# ca
n1 <- ggplot(heart, aes(ca, group = sex, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "flourosopy") +
  coord_flip() +
  scale_fill_manual(values = c("#386cb0","#fdb462"))+
  theme_classic2() +
  theme(legend.position='none')

ggarrange(g1, h1, i1, j1, k1, l1, m1, n1, 
          ncol = 2, nrow = 4,
          align = "v")
```

**From this first plot, it appears that this dataset contains more males patients with a higher proportion of heart disease compared to female patients.**

### B Visualization of variables per disease status
#### 1 Male patient

```{r}
heart <- heart2 %>%
    filter(sex == "male")
```

```{r}
# Male and Female count
a2 <- ggplot(heart, aes(x = target, fill = target)) +
  geom_bar(width = 0.5, position = 'dodge') + 
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# Age per gender
b2 <- ggplot(heart, aes(x= target, y = age, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) +
  ylim(0, 90) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# trestbps
c2 <- ggplot(heart, aes(x = target, y = trestbps, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "blood pressure (mmHg)") +
  ylim(0,250) +
 stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# chol
d2 <- ggplot(heart, aes(x = target, y = chol, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "cholestorol (mg/dl)") +
  ylim(0,500) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# oldpeak
e2 <- ggplot(heart, aes(x = target, y = oldpeak, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "ST depression") +
  ylim(0,10) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# thalach
f2 <- ggplot(heart, aes(x = target, y = thalach, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "Max. heart rate") +
  ylim(0,250) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

ggarrange(a2, b2, c2, d2, e2, f2, 
          ncol = 2, nrow = 3,
            align = "v")
```

**Male patients with heart disease are significantly older, have higher cholesterol level, and reduced maximum heart rate response to the thallium test. **

```{r}
# Disease status
g2 <- ggplot(heart, aes(x = target, fill = target)) +
  geom_bar(width = 0.5, position = 'dodge') + 
  labs(x = "") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# cp
h2 <- ggplot(heart, aes(cp, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "chest pain") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# restecg
i2 <- ggplot(heart, aes(restecg, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "rest. electrocardiographic") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# slope
j2 <- ggplot(heart, aes(slope, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "peak exercise ST") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# thal 
k2 <- ggplot(heart, aes(thal, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Thalium stress test") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# fbp
l2 <- ggplot(heart, aes(fbs, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Fasting blood sugar") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# exang
m2 <- ggplot(heart, aes(exang, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Exercise induced angina") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# ca
n2 <- ggplot(heart, aes(ca, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "flourosopy") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

ggarrange(g2, h2, i2, j2, k2, l2, m2, n2, 
          ncol = 2, nrow = 4,
          align = "v")
```

#### 2 Female patients

```{r}
heart <- heart2 %>%
    filter(sex == "female")
```

```{r}
# Male and Female count
a2 <- ggplot(heart, aes(x = target, fill = target)) +
  geom_bar(width = 0.5, position = 'dodge') + 
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# Age per gender
b2 <- ggplot(heart, aes(x= target, y = age, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) +
  ylim(0, 90) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# trestbps
c2 <- ggplot(heart, aes(x = target, y = trestbps, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "blood pressure (mmHg)") +
  ylim(0,250) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# chol
d2 <- ggplot(heart, aes(x = target, y = chol, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "cholestorol (mg/dl)") +
  ylim(0,500) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# oldpeak
e2 <- ggplot(heart, aes(x = target, y = oldpeak, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "ST depression") +
  ylim(0,10) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# thalach
f2 <- ggplot(heart, aes(x = target, y = thalach, fill = target)) +
  geom_violin(width = 0.5) +
  geom_boxplot(width = 0.2) + 
  labs(y = "Max. heart rate") +
  ylim(0,250) +
  stat_compare_means(aes(label = ..p.signif..), method = "t.test") +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

suppressWarnings(ggarrange(a2, b2, c2, d2, e2, f2, 
          ncol = 2, nrow = 3,
            align = "v"))
```

**There is less woman with heart disease on this data set. 
Women with heart disease have a significantly higher resting blood presure contrary to male with heart disease.
Similarly to men, women with heart disease have a lower maximum heart rate in response to the thallium test. 
**

```{r}
# Disease status
g2 <- ggplot(heart, aes(x = target, fill = target)) +
  geom_bar(width = 0.5, position = 'dodge') + 
  labs(x = "") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# cp
h2 <- ggplot(heart, aes(cp, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "chest pain") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# restecg
i2 <- ggplot(heart, aes(restecg, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "rest. electrocardiographic") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# slope
j2 <- ggplot(heart, aes(slope, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "peak exercise ST") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# thal 
k2 <- ggplot(heart, aes(thal, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Thalium stress test") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# fbp
l2 <- ggplot(heart, aes(fbs, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Fasting blood sugar") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# exang
m2 <- ggplot(heart, aes(exang, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "Exercise induced angina") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

# ca
n2 <- ggplot(heart, aes(ca, group = target, fill = target)) +
  geom_bar(position = "dodge") +
  labs(x = "", y = "flourosopy") +
  coord_flip() +
  scale_fill_manual(values = c("#7fc97f","#ef3b2c"))+
  theme_classic2() +
  theme(legend.position='none')

ggarrange(g2, h2, i2, j2, k2, l2, m2, n2, 
          ncol = 2, nrow = 4,
          align = "v")
```

## Results
### Setting up the models

```{r}
heart <- copy %>%
  filter(
    thal != 0 & ca != 4 # remove values correspondind to NA in original dataset
    )
  
```

```{r}
# ggcorr(heart, palette = "RdBu")
GGally::ggcorr(heart, geom = "circle")
```

```{r}
select2 <- heart %>%
  dplyr::select(
    target,
    slope,
    thalach,
    restecg,
    cp
  )
```

```{r}
ggcorr(select2, geom = "circle")
```

```{r}
ggpairs(heart)
```

```{r}
ggpairs(select2)
```

From the correlation study it seems that the parameters    
* cp  
* restecg  
* thalach  
* slope   
are the most usefull to predict the risk of heart disease  

From the EDA anlysis it semms that  
* age  
* sex  
* cholesterol  
* restecg   
are also usefull  

For prediction the following variables seems the most usefull  
* age  
* sex  
* cholesterol  
* restecg  
* cp  
* thalach  
* slope  

## VI Machine Learning: classification model with rpart and random forest packages
1. Select the columns usefull for prediction according to the EDA analysis.  
2. Separate the data set in a train and test subsets.  
3. Build a classification tree model with rpart.  
4. Print model accuracy and descision tree.  

### A Use select columns for classification
```{r}
# glimpse(heart)

heart_select <- heart %>%
  dplyr::select( #because of conflict between MASS and dplyr select need to use dplyr::select
    target,
    age,
    sex,
    chol,
    restecg,
    cp,
    thalach,
    slope
  )
```

```{r}
heart_select$target <- factor(heart_select$target) # Define target as a factor. rpart classification would not work otherwise.

accuracy <- 0

# Build a simple classification desicion tree with rpart. Run the model until the accuracy reach the selected minimum.
while(accuracy <= 0.88) {
  split_values <- sample.split(heart_select$target, SplitRatio = 0.65)
  train_set <- subset(heart_select, split_values == T)
  test_set <- subset(heart_select, split_values == F)
  mod_class <- rpart(target~. , data=train_set)
  result_class <- predict(mod_class, test_set, type = "class")
  table <- table(test_set$target, result_class)
  accuracy <- (table["0","0"] + table["1","1"])/sum(table)
  # cat("accuracy = ", round(accuracy, digits = 2)*100, "%")
}
```

Print model accuracy.  
According to parameters the model should be at least 88% accurate.  
```{r}
cat("Model accuracy", round(accuracy, digits = 2)*100, "%")  
```

Print the desicion tree.  
```{r}
# par(mfrow = c(1,2), xpd = NA) # otherwise on some devices the text is clipped
fancyRpartPlot(mod_class, , caption = NULL)
# plot(mod_class)
# text(mod_class, use.n = TRUE)
```

### B Use the full dataset for classification
```{r}
copy2 <- heart
heart$target <- factor(heart$target)
accuracy <- 0

# Build a simple classification desicion tree with rpart. Run the model until the accuracy reach the selected minimum.
while(accuracy <= 0.88) {
  split_values <- sample.split(heart_select$target, SplitRatio = 0.65)
  train_set <- subset(heart, split_values == T)
  test_set <- subset(heart, split_values == F)
  mod_class <- rpart(target~. , data=train_set)
  result_class <- predict(mod_class, test_set, type = "class")
  table <- table(test_set$target, result_class)
  accuracy <- (table["0","0"] + table["1","1"])/sum(table)
  # cat("accuracy = ", round(accuracy, digits = 2)*100, "%")
}
```

Print model accuracy.  
According to parameters the model should be at least 88% accurate.  
```{r}
cat("Model accuracy", round(accuracy, digits = 2)*100, "%")  
```

Print the desicion tree.  
```{r}
# par(mfrow = c(1,2), xpd = NA) # otherwise on some devices the text is clipped
fancyRpartPlot(mod_class, , caption = NULL)
# plot(mod_class)
# text(mod_class, use.n = TRUE)
```

### C Prediction on selected column with random forest
```{r}
set.seed(103)
train <- sample(nrow(heart_select), 0.7*nrow(heart_select), replace = FALSE)
TrainSet <- heart_select[train,]
ValidSet <- heart_select[-train,]
summary(TrainSet)
summary(ValidSet)
```

```{r}
# Create a Random Forest model with default parameters
model1 <- randomForest(target ~ ., data = TrainSet,  ntree = 1000, mtry = 1, importance = TRUE)
model1
```

```{r}
# Predicting on train set
predTrain <- predict(model1, TrainSet, type = "class")
# Checking classification accuracy
table(predTrain, TrainSet$target) 
```

```{r}
# Predicting on Validation set
predValid <- predict(model1, ValidSet, type = "class")
# Checking classification accuracy
mean(predValid == ValidSet$target)                    
table(predValid,ValidSet$target)
```

```{r}
# To check important variables
importance(model1)        
varImpPlot(model1)  
```


### D Use the full dataset for classification with random forest
```{r}
set.seed(103)
train <- sample(nrow(heart), 0.7*nrow(heart_select), replace = FALSE)
TrainSet <- heart[train,]
ValidSet <- heart[-train,]
summary(TrainSet)
summary(ValidSet)
```

```{r}
# Create a Random Forest model with default parameters
model2 <- randomForest(target ~ ., data = TrainSet,  ntree = 1000, mtry = 2, importance = TRUE)
model2
```

```{r}
# Predicting on train set
predTrain <- predict(model2, TrainSet, type = "class")
# Checking classification accuracy
table(predTrain, TrainSet$target) 
```

```{r}
# Predicting on Validation set
predValid <- predict(model2, ValidSet, type = "class")
# Checking classification accuracy
mean(predValid == ValidSet$target)                    
table(predValid,ValidSet$target)
```

```{r}
# To check important variables
importance(model2)        
varImpPlot(model2)  
```


## Conclusion
We set out to use the UCI data set on Heart Disease to create a model that could correctly predict Heart Disease diagnoses. We set a goal of achieving an validation set accuracy score of 0.85. We started by downloading the UCI data set on Heart Disease. We then cleaned the data set and prepared it for analysis. We split the data set into training and test sets.  We found that having a heart defect, the number of major vessels that were working, and the type of chest pain we the most important factors in determining if you have heart disease or not, using this data set. We achieved our goal of creating a model with a validation set accuracy score of 0.88 with the Random Forest model!

### Limitations
For me the biggest limitation in this project is the size of the data set. With only 303 observations this is a very small sample size. The other limitation is the data within the data set. 14 features is enough to achieve a high prediction accuracy, as we proved, but I think with more features we could achieve a score over 91%.

### Future Work
For the future, I would be curious to see how these algorithms preform on a much larger data set of data of our country. Along with a data set that has more features such as: height, weight, if parents had heart disease, use of drugs and alcohol, exercise amount, etc. I would be curious to see which algorithms preform better and if any preform worse. One final thing that I would include is adding more algorithms to this project. These 9 algorithms are not the only algorithms that work well with classification and they may produce a higher validation set accuracy score.

## References
Data transformation  
https://archive.ics.uci.edu/ml/datasets/Heart+Disease  
https://lucdemortier.github.io/projects/3_mcnulty  
https://www.kaggle.com/ronitf/heart-disease-uci/discussion/105877  

Kaggles notebooks:  
R notebooks:  
https://www.kaggle.com/ekrembayar/heart-disease-uci-eda-models-with-r  
https://www.kaggle.com/joemenifee/heart-disease-uci-data-exploratory  

Data Processing  
http://www.cookbook-r.com/
https://bookdown.org/rdpeng/exdata/managing-data-frames-with-the-dplyr-package.html#data-frames 
https://rpkgs.datanovia.com/ggpubr/reference/stat_compare_means.html  
https://towardsdatascience.com/simple-fast-exploratory-data-analysis-in-r-with-dataexplorer-package-e055348d9619  
http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/  
https://www.r-graph-gallery.com/267-reorder-a-variable-in-ggplot2  
for categorical variable  
https://www.analyticsvidhya.com/blog/2015/11/easy-methods-deal-categorical-variables-predictive-modeling/  
for correlations  
http://www.sthda.com/english/wiki/ggally-r-package-extension-to-ggplot2-for-correlation-matrix-and-survival-plots-r-software-and-data-visualization  
https://www.kaggle.com/code/jarredpriester/heart-disease-predictions-using-a-ml-ensemble-in-r/notebook
https://www.kaggle.com/code/wguesdon/predicting-heart-disease-risk-with-random-forest/script
